Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-22 paper/main.tex
finished GLMM methods section

In each version of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and
separately for each of the three quizzes (i.e., column in
Fig.~\ref{fig:predictions}), we then fit a generalized linear mixed model
(GLMM) with a logistic link function to the set of knowledge estimates for all
questions that participants answered on the given quiz. We implemented these
models in \texttt{R} using the \texttt{lme4} package~\citep{BateEtal15a} and
fit them following guidance from~\citet{BateEtal15b} and \citet{MatuEtal17}.
Specifically, we initially fit each model with the maximal random effects
structure afforded by our design, which we identified as:
\[
    \mathtt{accuracy \sim knowledge + (knowledge\ \vert\ participant) + (knowledge\ \vert\ question)}
\]
where ``\texttt{accuracy}'' is a binary value indicating whether each target
question was answered correctly or incorrectly, ``\texttt{knowledge}'' is
estimated knowledge at each target question's embedding coordinate,
``\texttt{participant}'' is a unique identifier assigned to each participant,
and ``\texttt{question}'' is a unique identifier assigned to each quiz
question. For models we fit using knowledge estimates for target questions
about multiple content areas (i.e., in the ``All questions'' version of the
analysis), we also included an additional random effect term,
$\mathtt{(knowledge\ \vert\ lecture)}$, where ``\texttt{lecture}'' is a
categorical value denoting whether the target question was about \textit{Four
Fundamental Forces}, \textit{Birth of Stars}, or general physics knowledge.
Note that with our coding scheme, identifiers for each \texttt{question} are
implicitly nested within levels of \texttt{lecture} and do not require explicit
nesting in our model formula.

% We then iteratively removed random effects from the maximal model until it
% successfully converged with a full rank (i.e., non-singular) random effects
% variance-covariance matrix.

@@

To assess the predictive value of our knowledge estimates, we compared each
GLMM's ability to discriminate between correctly and incorrectly answered
questions to that of an analogous model that did not consider estimated
knowledge. Specifically, we used the same sets of observations with which we
fit each ``full'' model to fit a second ``null'' model, with the formula:
\[
    \mathtt{accuracy \sim (1\ \vert\ participant) + (1\ \vert\ question)}
\]
where the terms are as defined above.

%In order to assess the predictive value of the knowledge estimates, we then fit a second set of ``null'' models to the same sets of observations used to fit our full GLMMs. These 
%
%
%used the same sets of observations used to fit these GLMMs to fit a second set of ``null'' models. 
%
%
%
%Next, in order to assess the predictive value of the knowledge estimates 
%
%
%
%In order to assess the predictive value of the knowledge estimates we used to fit each GLMM, we then fit a \textit{second} model to the same data as each of the 15 
%
%In order to assess whether the knowledge estimates we used to fit each GLMM could reliably predict participants' success on held-out questions, we then fit a second GLMM to the observations

%$$
%\begin{aligned}
%  \operatorname{accuracy}_{i}  &\sim \operatorname{Binomial}(n = 1, \operatorname{prob}_{\operatorname{accuracy} = \operatorname{correct}} = \widehat{P}) \\
%    \log\left[\frac{\hat{P}}{1 - \hat{P}} \right] &=\alpha_{j[i],k[i]} \\
%\left(
%  \begin{array}{c}
%    \begin{aligned}
%      &\alpha_{j} \\
%      &\gamma_{1j}
%    \end{aligned}
%  \end{array}
%\right)
%  &\sim N \left(
%\left(
%  \begin{array}{c}
%    \begin{aligned}
%      &\gamma_{0}^{\alpha} + \gamma_{1k[i]}^{\alpha}(\operatorname{knowledge}) \\
%      &\mu_{\gamma_{1j}}
%    \end{aligned}
%  \end{array}
%\right)
%,
%\left(
%  \begin{array}{cc}
%     \sigma^2_{\alpha_{j}} & \rho_{\alpha_{j}\gamma_{1j}} \\
%     \rho_{\gamma_{1j}\alpha_{j}} & \sigma^2_{\gamma_{1j}}
%  \end{array}
%\right)
% \right)
%    \text{, for participant j = 1,} \dots \text{,J} \\
%\left(
%  \begin{array}{c}
%    \begin{aligned}
%      &\alpha_{k} \\
%      &\gamma_{1k}
%    \end{aligned}
%  \end{array}
%\right)
%  &\sim N \left(
%\left(
%  \begin{array}{c}
%    \begin{aligned}
%      &\mu_{\alpha_{k}} \\
%      &\mu_{\gamma_{1k}}
%    \end{aligned}
%  \end{array}
%\right)
%,
%\left(
%  \begin{array}{cc}
%     \sigma^2_{\alpha_{k}} & \rho_{\alpha_{k}\gamma_{1k}} \\
%     \rho_{\gamma_{1k}\alpha_{k}} & \sigma^2_{\gamma_{1k}}
%  \end{array}
%\right)
% \right)
%    \text{, for question k = 1,} \dots \text{,K}
%\end{aligned}
%$$

>>>>

In each version of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and
separately for each of the three quizzes (i.e., column in
Fig.~\ref{fig:predictions}), we then fit a generalized linear mixed model
(GLMM) with a logistic link function to the set of knowledge estimates for all
questions that participants answered on the given quiz. We implemented these
models in \texttt{R} using the \texttt{lme4} package~\citep{BateEtal15a} and
fit them following guidance from~\citet{BateEtal15b} and \citet{MatuEtal17}.
Specifically, we initially fit each model with the maximal random effects
structure afforded by our design, which we identified as:
\[
    \mathtt{accuracy \sim knowledge + (knowledge\ \vert\ participant) + (knowledge\ \vert\ question)}
\]
where ``\texttt{accuracy}'' is a binary value indicating whether each target
question was answered correctly or incorrectly, ``\texttt{knowledge}'' is
estimated knowledge at each target question's embedding coordinate,
``\texttt{participant}'' is a unique identifier assigned to each participant,
and ``\texttt{question}'' is a unique identifier assigned to each quiz
question. For models we fit using knowledge estimates for target questions
about multiple content areas (i.e., in the ``All questions'' version of the
analysis), we also included an additional random effect term,
$\mathtt{(knowledge\ \vert\ lecture)}$, where ``\texttt{lecture}'' is a
categorical value denoting whether the target question was about \textit{Four
Fundamental Forces}, \textit{Birth of Stars}, or general physics knowledge.
Note that with our coding scheme, identifiers for each \texttt{question} are
implicitly nested within levels of \texttt{lecture} and do not require explicit
nesting in our model formula. We then iteratively removed random effects from 
the maximal model until it successfully converged with a full rank (i.e., non-singular) 
random effects variance-covariance matrix.

@@

To assess the predictive value of our knowledge estimates, we compared each
GLMM's ability to discriminate between correctly and incorrectly answered
questions to that of an analogous model that did \textit{not} consider estimated
knowledge. Specifically, we used the same sets of observations with which we
fit each ``full'' model to fit a second ``null'' model, with the formula:
\[
    \mathtt{accuracy \sim (1\ \vert\ participant) + (1\ \vert\ question)}
\]
where ``\texttt{accuracy}'', ``\texttt{participant}'', and ``\texttt{question}'' are as defined above. 
As with our full models, the null models we fit for the “All questions” version of the analysis for each quiz contained an additional term, $\mathtt{(1\ \vert\ lecture)}$, where ``\texttt{lecture}'' are as defined above. 
We then compared each full model to its reduced (null) equivalent using a likelihood-ratio test (LRT). 
Because the typical asymptotic $\chi^2_d$ approximation of the null distribution for the LRT statistic ($\lambda_{LR}$) is anti-conservative for models that differ in their random slope terms~\citep{GoldSimo00,ScheEtal08b,SnijBosk11}, we computed $p$-values for these tests using a parametric bootstrapping procedure~\citep{HaleHojs14}. 
For each of 1,000 bootstraps, we used the fitted null model to simulate a sample of observations of equal size to our original sample.
We then re-fit both the null and full models to this simulated sample and compared them via an LRT.
This yielded a distribution of $\lambda_{LR}$ statistics we may expect to observe under our null hypothesis.
Following~\citep{DaviHink97,NortEtal02}, we computed a corrected $p$-value for our observed $\lambda_{LR}$ as $\frac{r + 1}{n + 1}$, where $r$ is the number of simulated model comparisons that yielded a $\lambda_{LR}$ greater than or equal to our observed value and $n$ is the number of simulations we ran (1,000).
